---
title: "Supplementary Material: Retracted articles use less free and open source software and cite it worse"
author: "David Schindler and Frank Krüger"
format: pdf
toc: false
code-line-numbers: true
editor: visual
execute: 
  warning: false
bibliography: references.bib
suppress-bibliography: true
---

# Software Usage and Citation in Retracted Articles

This document is created as a literate data analysis where we provide descriptions, implemented code, and the results in one publication. It summarizes all data analyses concerning the scientific software landscape and software citation habits in retracted articles.

```{r}
library(tidyverse)
library(magrittr)
library(patchwork)
library(effectsize)

theme_set(theme_bw() +   
            theme(legend.position = 'top', 
                  strip.background = element_rect(fill="#E5E5E5"), 
                  plot.margin = unit(c(0,2,0,0), 'mm'),
                  legend.margin=margin(0,0,0,0),
                  legend.box.margin=margin(0,0,-5,0),
                  plot.caption.position = "plot",
                  plot.caption = element_text(hjust = 0),
                  plot.tag.position = "bottomleft")
)
```

## Loading Data

We load anonymized information on software in retracted and control articles. This includes identified software and additional information such as version, developer, and software type. The information was aggregated by information extraction from full-text articles. Control articles were selected by Coarsened Exact Matching. The published data which is loaded here is based on data provided by Retraction Watch and published with their permission. The original data is available from Retraction Watch. Below we describe all information sources used to create the data.

```{r}
df <- read_csv("software_in_retracted_and_control_articles.csv")
```

### RW Article Retractions

We obtained the RW database on article retractions as of January 6th, 2022. We use information on retraction, the reason for retraction, journal, and DOI. We perform full-text analyses of these articles and can, therefore, only include articles for which plain full-text is available. This excludes a large number of articles, but the remaining sample size is still sufficient to perform a large-scale analysis, as shown by prior studies investigating article retractions, facing similar issues, and working on comparable sample sizes [@peng2022dynamics].

### S2ORC

The full-text data is sampled from [S2ORC](https://github.com/allenai/s2orc/) [@lo-etal-2020-s2orc], currently the largest source of available scientific publications in plain text format. The information on software contained in articles is extracted from these full-texts. Further, metadata on scientific domain, year, and journal from S2ORC was utilized.

### Summarized Retraction Reasons

There are several retraction reasons considered by Retraction Watch. An overview is available at [reasons](https://retractionwatch.com/retraction-watch-database-user-guide/retraction-watch-database-user-guide-appendix-b-reasons/). However, these are too fine-grained to allow meaningful analyses of the given data set. Therefore, following prior work [@ribeiro2018retractions], we manually summarized reasons into broader categories. The top-level reasons we consider are:

1.  Error: Honest errors in investigations that can occur due to multiple reasons.
2.  Investigation: Investigations into publications performed by different parties, for instance, by the publisher or an institution.
3.  Plagiarism: Cases in which prior work was used without correctly indicating the source
4.  SelfPlagiarism: Duplication of one's own prior work without indicating the source.
5.  Misconduct: Cases in which a scientific misconduct was performed by the authors.
6.  PaperMill: Articles that were automatically generated with paper mill techniques (and are not supported by actual research).
7.  other: Category that matches all unmatched reasons. There are several unspecific reasons that are not related to others, e.g., "rogue editor".

```{r}
reasons <- c('Error', 'Investigation', 'Plagiarism', 'SelfPlagiarism', 
             'Misconduct', 'PaperMill', 'other')

read_csv("retraction_reasons.csv") %>%
  mutate(TopReason = factor(TopReason, levels = reasons)) -> reasons_print
split(reasons_print$Reason, reasons_print$TopReason) 
```

### Control Articles

We select a set of control articles by Coarsened Exact Matching (CEM) [@iacus2012causal]. Three article attributes are controlled that have a proven influence on software usage and citation habits [@schindler2022role]:

1.  **Publication date**: coarsened to *year.* The generally observed trend is that software usage increases over time.

2.  **Scientific domain**: matched *exactly*. Specific domains were observed to exhibit higher/lower software usage/citation quality. Domain order for multidisciplinary work is retained: \[Computer Science, Biology\] is different from \[Biology, Computer Science\]

3.  **Journal Rank:** coarsened to *percentiles*. Higher journal rank has been associated with more formal software citations attributed to more comprehensive journal policies.

Year and domain are determined from Retraction Watch and S2ORC metadata, while the journal rank is based on the Scimago Journal Rank (SJR). [Scimago](https://www.scimagojr.com/) offers publicly available information on journal rank on a yearly basis, which we gathered directly from the Website. The journals information for articles is added by matching Retraction Watch and S2ORC journal information with the Scimago Journal entries.

### Software Information Enrichment

Here, we load manually annotated information on software generated during information enrichment:

1.  software availability: free and commercial
2.  source code availability: open-source and closed-source
3.  whether software is statistical software

```{r}
software_enrichment <- read_csv('software_enrichment.csv', na = 'na') %>%
  drop_na(free)
```

```{r}
n_free <- nrow(filter(software_enrichment, free==1))
n_free_and_open_source <- nrow(filter(
  software_enrichment, free==1 & source==1))
n_free_and_not_open_source <- nrow(filter(
  software_enrichment, free==1 & source==0))
n_open <- nrow(filter(software_enrichment, source==1))
paste0(
  round(n_free_and_open_source/n_free, digits=2), 
  "% of free software are also open source, ", 
  round(n_free_and_not_open_source/n_free, digits=2), 
  "% are not open source.")
paste0(
  round(n_free_and_open_source/n_open, digits=2), 
  "% of open source software are also free.")
```

## Data Corrections

We manually correct disambiguation errors that were identified during information enrichment. There are two types of cases: false positive disambiguation, where software names were linked even so they refer to different software, and false negative disambiguation where software names were not linked even so they refer to the same software. Overall, there were 12 false negative cases of software groups and 10 false positive errors of software groups. Groups refer to larger errors where names that appear multiple times are added to other groups that also appear that often. Additionally, there were 14 cases of false negatives where single occurrences were linked to a group, which we consider a small error. All individual errors are corrected here:

```{r}
df %<>%
  mutate(Software_ID = ifelse(Software_Name == 'Image J', 
                     43391, 
                     Software_ID)) %>% 
  # false negative matching - big error
  mutate(Software_Name = ifelse(Software_Name == 'Image J', 
                       "ImageJ", 
                       Software_Name)) %>%
  mutate(Software_ID = ifelse(grepl("scion", Software_String, 
                                    ignore.case = TRUE), 
                     43514, 
                     Software_ID)) %>% 
  # false postive linking - big error
  mutate(Software_Name = ifelse(grepl("scion", Software_String, 
                                      ignore.case = TRUE), 
                       "Scion Image", 
                       Software_Name)) %>%
  mutate(Software_ID = ifelse(grepl("^limma$|^limma ", 
                           Software_String, ignore.case = TRUE), 
                     43771, 
                     Software_ID)) %>% 
  # false negative linking - big error
  mutate(Software_Name = ifelse(grepl("^limma$|^limma ", 
                             Software_String, ignore.case = TRUE), 
                       "limma", 
                       Software_Name)) %>%
  mutate(Software_ID = ifelse(grepl("coot", Software_String, 
                                    ignore.case = TRUE), 
                     43895, 
                     Software_ID)) %>% 
  # false negative linking - big error
  mutate(Software_Name = ifelse(grepl("coot", Software_String, 
                                      ignore.case = TRUE), 
                       "COOT", 
                       Software_Name)) %>%
  mutate(Software_ID = ifelse(grepl("pasw", Software_String, 
                                    ignore.case = TRUE), 
                     43381, 
                     Software_ID)) %>% 
  # false negative linking - big error
  mutate(Software_Name = ifelse(grepl("pasw", Software_String, 
                                      ignore.case = TRUE), 
                       "SPSS", 
                       Software_Name)) %>%
  mutate(Software_ID = ifelse(grepl("fastx", Software_String, 
                                    ignore.case = TRUE), 
                     46206, 
                     Software_ID)) %>% 
  # false positive and false negative linking - big error
  mutate(Software_Name = ifelse(grepl("fastx", Software_String, 
                                      ignore.case = TRUE), 
                       "FASTX - Toolkit", 
                       Software_Name)) %>%
  mutate(Software_ID = ifelse(grepl("tblastn", Software_String, 
                                    ignore.case = TRUE), 
                     45848, 
                     Software_ID)) %>% 
  # false negative linking - big error
  mutate(Software_Name = ifelse(grepl("tblastn", Software_String, 
                                      ignore.case = TRUE), 
                       "tblastn", 
                       Software_Name)) %>%
  mutate(Software_ID = ifelse(grepl("macintosh", Software_String, 
                                    ignore.case = TRUE), 
                     43695, 
                     Software_ID)) %>% 
  # false negative linking - big error
  mutate(Software_Name = ifelse(grepl("macintosh", Software_String, 
                                      ignore.case = TRUE), 
                       "Mac", 
                       Software_Name)) %>%
  mutate(Software_ID = ifelse(grepl("Significance Analysis of Microarrays", 
                           Software_String, ignore.case = TRUE), 
                     44860, 
                     Software_ID)) %>% 
  # false negative linking - big error
  mutate(Software_Name = ifelse(grepl("Significance Analysis of Microarrays", 
                             Software_String, ignore.case = TRUE), 
                       "SAM", 
                       Software_Name)) %>%
  mutate(Software_ID = ifelse(grepl("NetworkX", Software_String,
                                    ignore.case = TRUE), 
                     54000, 
                     Software_ID)) %>% 
  # false positive linking - big error
  mutate(Software_Name = ifelse(grepl("NetworkX", Software_String, 
                                      ignore.case = TRUE), 
                       "NetworkX", 
                       Software_Name)) %>%
  mutate(Software_ID = ifelse(grepl("microarray suite", Software_String, 
                                    ignore.case = TRUE), 
                     45171, 
                     Software_ID)) %>% 
  # false negative linking - big error
  mutate(Software_Name = ifelse(grepl("microarray suite", 
                             Software_String, ignore.case = TRUE), 
                      "MAS", 
                      Software_Name)) %>%
  mutate(Software_ID = ifelse(grepl("Statistical Parametric Mapping", 
                           Software_String, ignore.case = TRUE), 
                     44213, 
                     Software_ID)) %>% 
  # false negative linking - big error
  mutate(Software_Name = ifelse(grepl("Statistical Parametric Mapping", 
                             Software_String, ignore.case = TRUE), 
                       "SPM", 
                       Software_Name)) %>%
  mutate(Software_ID = ifelse(Software_String == 'IPA' | 
                                Software_String == 'IPA TM', 
                     44597, 
                     Software_ID)) %>% 
  # false negative linking - big error
  mutate(Software_Name = ifelse(Software_String == 'IPA' | 
                                  Software_String == 'IPA TM', 
                       "Ingenuity Pathway Analysis", 
                       Software_Name)) %>%
  mutate(Software_ID = ifelse(Software_String == "Ingenuity" | 
                       grepl("Ingenuity", Software_String, 
                             ignore.case = TRUE) & 
                       (grepl("pathway", Software_String, 
                              ignore.case = TRUE) | 
                          grepl("ipa", Software_String, 
                                ignore.case = TRUE) | 
                          grepl("system", Software_String, 
                                ignore.case = TRUE)), 
                     44597, 
                     Software_ID)) %>% 
  # false negative linking - big error
  mutate(Software_Name = ifelse(Software_String == "Ingenuity" | 
                         grepl("Ingenuity", Software_String, 
                               ignore.case = TRUE) & 
                         (grepl("pathway", Software_String, 
                                ignore.case = TRUE) | 
                            grepl("ipa", Software_String, 
                                  ignore.case = TRUE) | 
                            grepl("system", Software_String, 
                                  ignore.case = TRUE)), 
                       "Ingenuity Pathway Analysis", 
                       Software_Name)) %>%
  mutate(Software_ID = ifelse(grepl('^statistics$', Software_String, 
                                    ignore.case = TRUE), 
         43381, 
         Software_ID)) %>% 
  # false positive linking - big error
  mutate(Software_Name = ifelse(grepl('^statistics$', Software_String, 
                                      ignore.case = TRUE), 
         "SPSS", 
         Software_Name)) %>%
  mutate(Software_ID = ifelse(grepl('^gcos$', Software_String, 
                                    ignore.case = TRUE), 
         55013, 
         Software_ID)) %>% 
  # false positive linking - big error
  mutate(Software_Name = ifelse(grepl('^gcos$', Software_String, 
                                      ignore.case = TRUE), 
         "GCOS", 
         Software_Name)) %>%
  mutate(Software_ID = ifelse(grepl('^chrome$', Software_String, 
                                    ignore.case = TRUE), 
         45207, 
         Software_ID)) %>% 
  # false positive linking - big error
  mutate(Software_Name = ifelse(grepl('^chrome$', Software_String, 
                                      ignore.case = TRUE), 
         "Google Chrome", 
         Software_Name)) %>%
  mutate(Software_ID = ifelse(grepl('^primer ?premier$', Software_String, 
                                    ignore.case = TRUE), 
         55012, 
         Software_ID)) %>% 
  # false positive linking - big error
  mutate(Software_Name = ifelse(grepl('^primer ?premier$', Software_String, 
                                      ignore.case = TRUE), 
         "Primer Premier", 
         Software_Name)) %>%
  mutate(Software_ID = ifelse(grepl('^mr ?modeltest$', Software_String, 
                                    ignore.case = TRUE), 
         43157, 
         Software_ID)) %>% 
  # false positive linking - big error
  mutate(Software_Name = ifelse(grepl('^mr ?modeltest$', Software_String, 
                                      ignore.case = TRUE), 
         "MrModelTest", 
         Software_Name)) %>%
  mutate(Software_ID = ifelse(grepl('^unix$', Software_String, 
                                    ignore.case = TRUE), 
         55010, 
         Software_ID)) %>% 
  # false positive linking - big error
  mutate(Software_Name = ifelse(grepl('^unix$', Software_String, 
                                      ignore.case = TRUE), 
         "UNIX", 
         Software_Name)) %>%
  mutate(Software_ID = ifelse(grepl('Java ?Tree ?View', Software_String, 
                                    ignore.case = TRUE), 
         55008, 
         Software_ID)) %>% 
  # false positive linking - big error
  mutate(Software_Name = ifelse(grepl('Java ?Tree ?View', Software_String, 
                                      ignore.case = TRUE), 
         "Java TreeView", 
         Software_Name)) %>%
  mutate(Software_ID = ifelse(Software_String == 'MIRA' | 
                                Software_String == 'Mira', 
                     55001, 
                     Software_ID)) %>% 
  # false positive linking - big error
  mutate(Software_Name = ifelse(Software_String == 'MIRA' | 
                                  Software_String == 'Mira', 
                       "MIRA", 
                       Software_Name)) %>%
  mutate(Software_ID = ifelse(grepl('^statistical$', Software_String, 
                                    ignore.case = TRUE), 
         43381, 
         Software_ID)) %>% 
  # false positive linking - small error
  mutate(Software_Name = ifelse(grepl('^statistical$', Software_String, 
                                      ignore.case = TRUE), 
         "SPSS", 
         Software_Name)) %>%
  mutate(Software_ID = ifelse(grepl('m ?- ?plus', Software_String, 
                                    ignore.case = TRUE), 
         43920, 
         Software_ID)) %>% 
  # false positive linking - small error
  mutate(Software_Name = ifelse(grepl('m ?- ?plus', Software_String, 
                                      ignore.case = TRUE), 
         "Mplus", 
         Software_Name)) %>%
  mutate(Software_ID = ifelse(grepl('treee?dit', Software_String, 
                                    ignore.case = TRUE), 
         55014, 
         Software_ID)) %>% 
  # false positive linking - small error
  mutate(Software_Name = ifelse(grepl('treee?dit', Software_String, 
                                      ignore.case = TRUE), 
         "TREEEDIT", 
         Software_Name)) %>%
  mutate(Software_ID = ifelse(grepl('redhat', Software_String, 
                                    ignore.case = TRUE), 
         19693, 
         Software_ID)) %>% 
  # false positive linking - small error
  mutate(Software_Name = ifelse(grepl('redhat', Software_String, 
                                      ignore.case = TRUE), 
         "RedHat", 
         Software_Name)) %>%
  mutate(Software_ID = ifelse(grepl('^direct ?x$', Software_String, 
                                    ignore.case = TRUE), 
         55011, 
         Software_ID)) %>% 
  # false positive linking - small error
  mutate(Software_Name = ifelse(grepl('^direct ?x$', Software_String, 
                                      ignore.case = TRUE), 
         "DirectX", 
         Software_Name)) %>%
  mutate(Software_ID = ifelse(Software_String =='NET' | 
                       Software_String == 'Net' | 
                       Software_String == 'Net Framework', 
                     55009, 
                     Software_ID)) %>% 
  # false positive linking - small error
  mutate(Software_Name = ifelse(Software_String =='NET' | 
                         Software_String == 'Net' | 
                         Software_String == 'Net Framework', 
                       "NET", 
                       Software_Name)) %>%
  mutate(Software_ID = ifelse(Software_String == 'e' | 
                       Software_String == 'E - ' | 
                       Software_String == 'e1071' | 
                       Software_String == 'e1701' | 
                       Software_String == 'E5640', 
                     55007, 
                     Software_ID)) %>% 
  # false positive linking - small error
  mutate(Software_Name = ifelse(Software_String == 'e' | 
                         Software_String == 'E - ' | 
                         Software_String == 'e1071' |
                         Software_String == 'e1701' | 
                         Software_String == 'E5640', 
                       "E", 
                       Software_Name)) %>%
  mutate(Software_ID = ifelse(grepl('after effect', Software_String, 
                                    ignore.case = TRUE), 
         55006, 
         Software_ID)) %>% 
  # false positive linking - small error
  mutate(Software_Name = ifelse(grepl('after effect', Software_String, 
                                      ignore.case = TRUE), 
         "After Effects",
         Software_Name)) %>%
  mutate(Software_ID = ifelse(grepl('^avid$', Software_String, 
                                    ignore.case = TRUE), 
         55005, 
         Software_ID)) %>% 
  # false positive linking - small error
  mutate(Software_Name = ifelse(grepl('^avid$', Software_String, 
                                      ignore.case = TRUE), 
         "CAVID", 
         Software_Name)) %>%
  mutate(Software_ID = ifelse(Software_String == 'C50', 
                     55004, 
                     Software_ID)) %>% 
  # false positive linking - small error
  mutate(Software_Name = ifelse(Software_String == 'C50', 
                       "C50", 
                       Software_Name)) %>%
  mutate(Software_ID = ifelse(Software_String == 'C2000', 
                     55003, 
                     Software_ID)) %>% 
  # false positive linking - small error
  mutate(Software_Name = ifelse(Software_String == 'C2000', 
                       "C2000", 
                       Software_Name)) %>%
  mutate(Software_ID = ifelse(Software_String == 'C2000', 
                     55003, 
                     Software_ID)) %>% 
  # false positive linking - small error
  mutate(Software_Name = ifelse(Software_String == 'C2000', 
                       "C2000", 
                       Software_Name)) %>%
  mutate(Software_ID = ifelse(Software_String == 'c', 
                     43445, 
                     Software_ID)) %>% 
  # false positive linking - small error
  mutate(Software_Name = ifelse(Software_String == 'c', 
                       "C", 
                       Software_Name)) %>%
  mutate(Software_ID = ifelse(grepl('^creative suite$', Software_String, 
                                    ignore.case = TRUE), 
         55002, 
         Software_ID)) %>% 
  # false positive linking - small error
  mutate(Software_Name = ifelse(grepl('^creative suite$', Software_String, 
                                      ignore.case = TRUE), 
         "Creative Suite", 
         Software_Name))
```

Further, we remove 2 systematic extraction errors that were identified during information enrichment. Both are due to a specialized method being mistaken for software.

```{r}
df %>%
  group_by(Paper_ID) %>%
  filter(any(c('BLOSUM', 'B3LYP') %in% Software_Name)) %>% 
  select(Set_ID, Paper_ID, Retraction_Reason, Control_Sample_Origin, 
         Year, Scientific_Domain, Journal_Rank_Percentile) ->
  paper_ids_removed

df %<>%
  filter(! Software_Name %in% c('BLOSUM', 'B3LYP'))
  
unique(paper_ids_removed) %>%
  filter(! Paper_ID %in% df$Paper_ID) -> paper_ids_to_add

df <- bind_rows(paper_ids_to_add, df) %>% ungroup()
```

Last, we define a second dataframe that contains the information for analyses based on retraction reasons. As we have 10 corresponding sample articles for each retracted article we can generate a separate control set for each retraction reason that is equally distributed concerning the controlled variables. We use this dataframe for extended analyses.

```{r}
df %>%  
  select(Paper_ID, Retraction_Reason, Software_ID, Software_Name, 
         Version, Developer, Citation, URL) %>%
  mutate(URL=ifelse(is.na(URL), FALSE, TRUE)) %>%
  filter(Retraction_Reason != 'non-retracted') %>% 
  rename(OriginalReason=Retraction_Reason) %>%
  distinct() %>%
  mutate(set='retracted') ->
  retracted_papers

df %>%  
  select(Paper_ID, Retraction_Reason, Software_ID, Software_Name, 
         Version, Developer, Citation, URL, Control_Sample_Origin) %>%
  mutate(URL=ifelse(is.na(URL), FALSE, TRUE)) %>%
  inner_join(retracted_papers, by=c('Control_Sample_Origin'='Paper_ID')) %>%
  select(Paper_ID, Software_ID=Software_ID.x, Software_Name=Software_Name.x, 
         Version=Version.x, Developer=Developer.x,Citation=Citation.x, 
         URL=URL.x, OriginalReason) %>%
  mutate(set='non-retracted') %>%
  distinct() ->
  non_retracted_papers

df_reason_sampled <- rbind(retracted_papers, non_retracted_papers)
```

# Results

## Retraction Reasons

First, we are getting an overview of the reasons for article retraction and their frequency in the given data. We only look at the manually summarized, top-level reasons as there are too many different specific reasons for a meaningful analysis.

```{r, fig.height=2.5}
df %>% 
  filter(Set_ID=='retracted') %>%
  dplyr::select(Paper_ID, Retraction_Reason) %>%
  distinct() %>%
  group_by(Retraction_Reason) %>%
  count() %>%
  ungroup() %>%
  mutate(Retraction_Reason=reorder(Retraction_Reason, n)) %>%
  ggplot(aes(Retraction_Reason, n)) +
  geom_bar(stat='identity', fill='lightblue')  + 
  geom_text(aes(label=n)) +
  labs(x='Reason for Retraction', y='Number of Articles', 
       caption = 'Fig. S1: Number of articles corresponding to each retraction 
    reasons. Articles can be retracted due to more than one 
    reason.') + 
  scale_fill_brewer(type='qual', palette = 6) +
  coord_flip() +
  theme(plot.caption = element_text(size=8))
```

## Software Usage in Retracted Articles

Now, we perform the analyses on the software landscape and citation styles.

### Papers that Mention Software

We start with a basic analyses by looking at the relative number of articles that contain software.

#### Overall

We directly compare the relative numbers between sets.

```{r}
df %>%  
  dplyr::select(Set_ID, Paper_ID, Software_ID) %>%
  group_by(Set_ID, Paper_ID) %>%
  summarize(has_software=ifelse(is.na(Software_ID), 0, 1),
            .groups = "drop") %>%
  distinct() %>%
  group_by(Set_ID, has_software) %>%
  summarize(n=n()) %>%
  mutate(rel = n/sum(n)) %>%
  group_by(Set_ID) %>%
  mutate(num=n, n=sum(n)) %>%
  ungroup() %>%
  filter(has_software==1) %>%
  mutate(SEM=sqrt((rel * (1-rel))/n), 
         MoE = sqrt((rel * (1-rel))/n) * 1.96) %>%
  mutate(CIu = rel + MoE, CIl = rel - MoE) %>%
  select(Set_ID, rel, CIl, CIu) %>% 
  mutate(rel=rel*100, CIl=CIl*100, CIu=CIu*100) %>%
  mutate(across(where(is.numeric), round, 1))
```

We further include a McNemar test for the paired, dichotomous data to test if there is a difference in the amount of articles mentioning software between retracted and control articles. The effect size is then calculated by an odds ratio between both groups.

```{r}
df %>%  
  dplyr::select(Set_ID, Paper_ID, Software_ID, Control_Sample_Origin) %>%
  group_by(Set_ID, Paper_ID, Control_Sample_Origin) %>%
  summarize(has_software=ifelse(is.na(Software_ID), 0, 1),
            .groups = "drop") %>%
  distinct() -> df_t

x <- split(df_t, df_t$Set_ID)

x$retracted %>%
  select(-Control_Sample_Origin) %>%
  inner_join(x$`non-retracted`, by=c("Paper_ID"="Control_Sample_Origin")) -> 
  df_tmp

df_tmp %>% group_by(has_software.x, has_software.y) %>% summarize(n = n()) -> 
  data

p <- matrix(
  rev(data$n), 
  nrow=2, 
  dimnames = list(
    "control" = c("software", "no-software"),  
    "retracted" = c("software", "no-software")))

mcnemar.test(p)

odds_ratio <- data$n[3] / data$n[2]
odds_ratio
```

#### Over Time

We compare the numbers from the first to the last analyzed year per set.

```{r}
df %>%  
  dplyr::select(Set_ID, Paper_ID, Software_ID, Year) %>%
  filter(Year %in% c(2000, 2019)) %>%
  group_by(Set_ID, Paper_ID, Year) %>%
  summarize(has_software=ifelse(is.na(Software_ID), 0, 1),
            .groups = "drop") %>%
  distinct() %>%
  group_by(Set_ID, has_software, Year) %>%
  summarize(n_has_software=n(),.groups='drop') %>%
  group_by(Set_ID, Year) %>%
  mutate(n_year=sum(n_has_software)) %>%
  mutate(rel = n_has_software/n_year) %>%
  ungroup() %>%
  filter(has_software==1) %>%
  mutate(SEM=sqrt((rel * (1-rel))/n_year), 
         MoE = sqrt((rel * (1-rel))/n_year) * 1.96) %>%
  mutate(CIu = rel + MoE, CIl = rel-MoE) %>%
  select(Set_ID, Year, rel, CIl, CIu)  %>% 
  mutate(rel=rel*100, CIl=CIl*100, CIu=CIu*100) %>%
  mutate(across(where(is.numeric), round, 1))
```

Here, we depict the course detailed over all years in the analyses.

```{r}
df %>%  
  dplyr::select(Set_ID, Paper_ID, Software_ID, Year) %>%
  group_by(Set_ID, Paper_ID, Year) %>%
  summarize(has_software=ifelse(is.na(Software_ID), 0, 1),
            .groups = "drop") %>%
  distinct() %>%
  group_by(Set_ID, has_software, Year) %>%
  summarize(n_has_software=n(),.groups='drop') %>%
  group_by(Set_ID, Year) %>%
  mutate(n_year=sum(n_has_software)) %>%
  mutate(rel = n_has_software/n_year) %>%
  ungroup() %>%
  filter(has_software==1) %>%
  mutate(SEM=sqrt((rel * (1-rel))/n_year), 
         MoE = sqrt((rel * (1-rel))/n_year) * 1.96) %>%
  mutate(CIu = rel + MoE, CIl = rel-MoE) %>%
  ggplot(aes(Year, rel)) + 
    geom_line(aes(color=Set_ID)) + 
    geom_ribbon(aes(ymin=CIl, ymax=CIu, fill=Set_ID), alpha=.2) + 
    labs(x='Year', y='Relative Amount of Articles', 
      caption = 'Fig. S2: Relative number of articles containing at least 
    one software over time for retracted and control articles. 
    95% CIs are indicated by lighter colored areas.') + 
    theme(plot.caption = element_text(size=8)) +
    scale_color_manual('Type of Article', 
                      values = c("#2b83ba", "#ff8585")) +
    scale_fill_manual('Type of Article', 
                      values = c("#2b83ba", "#ff8585")) 
```

Confidence intervals are especially large for retracted articles because the overall number of samples decreases due to the year-based split (especially for earlier years, where the fewest samples are available).

#### Per Retraction Reason

Further, we also look at the relative number concerning specific retraction reasons. Each retraction reason has its own control set, which is created by using the 10 control samples for each article per retraction reason.

```{r, fig.height=4.5, fig.width=8}
df_reason_sampled %>%
  group_by(Paper_ID, set, OriginalReason) %>%
  summarize(has_software=ifelse(is.na(Software_ID), 0, 1), 
            .groups = "drop") %>%
  distinct() %>%
  group_by(OriginalReason, set, has_software) %>%
  summarize(n=n()) %>% 
  mutate(rel = n/sum(n)) %>%
  group_by(OriginalReason) %>%
  mutate(n=sum(n)) %>%
  ungroup() %>%
  filter(has_software==1) %>%
  mutate(SEM=sqrt((rel * (1-rel))/n), 
         MoE = sqrt((rel * (1-rel))/n) * 1.96) %>%
  mutate(CIu = rel + MoE, CIl = rel-MoE)  %>%
  mutate(rel = rel*100, CIu=CIu*100, CIl=CIl*100) %>%
  mutate(set=ifelse(set=="non-retracted", "Control", 'Retracted')) %>%
  select(OriginalReason, set, rel, CIu, CIl) %>% print(., n=16) %>%
  mutate(plot="Amount of Articles with Software") %>%
  mutate(OriginalReason = factor(OriginalReason, levels=reasons)) %>%
  ggplot(aes(OriginalReason, rel)) + 
  geom_point(aes(color=set), position=position_dodge(width=.6)) + 
  geom_errorbar(aes(ymin=CIl, ymax=CIu, color=set), 
                position=position_dodge(width = .6), width=.5) +
  labs(x=element_blank(), y = "Relative Amount of Articles")  + 
  scale_y_continuous(breaks = c(40, 60, 80, 100), 
                     labels = c("40%", "60%", "80%", "100%")) +
  scale_color_manual('Type of Article', 
                     values = c("#2b83ba", "#ff8585")) + 
  facet_wrap(~ plot, nrow=2, scales='free_y') -> p1
  
p1 +
  labs(caption = 'Fig. S3: Relative amount of articles containing at least one 
    software compared between retracted and control set 
    divided by retraction reasons. A separate control set is 
    constructed for each retraction reasons by selecting the 
    ten corresponding articles for each retracted paper.') + 
  theme(plot.caption = element_text(size=10))

p1 <- p1 + theme(legend.position='top', 
        axis.text.x=element_blank())
```

### Number of Different Software

Next, we look at the average number of different software that is mentioned within articles that contain software.

#### Overall

First, the basic compare between sets.

```{r}
df %>%
  dplyr::select(Set_ID, Paper_ID, Software_ID) %>%
  group_by(Set_ID, Paper_ID) %>%
  summarize(n=ifelse(is.na(Software_ID), 0, n_distinct(Software_ID))) %>%
  filter(n > 0) %>%
  ungroup() %>%
  distinct() %>%
  group_by(Set_ID) %>%
  summarize(m=mean(n), sd=sd(n), num=n(), ) %>%
  mutate(CIl=m-(qt(p=.975, df=num-1)*(sd/sqrt(num))), 
         CIu = m+(qt(p=.975, df=num-1)*(sd/sqrt(num))))
```

We further include a two-sample t-test to test if there is a difference in the number of software provided between retracted and control articles. An unpaired t-test is selected as data is not exactly paired because articles without software are removed for this test and we are considering a quantitative variable with the number of software. The effect size is calculated by using Cohen's d.

```{r}
df %>%
  dplyr::select(Set_ID, Paper_ID, Software_ID, Control_Sample_Origin) %>%
  group_by(Set_ID, Paper_ID, Control_Sample_Origin) %>%
  summarize(n=ifelse(
    is.na(Software_ID), 
    0, 
    n_distinct(Software_ID)), 
    .groups = 'drop') %>%
  filter(n > 0) %>%
  distinct() -> df_t

x <- split(df_t, df_t$Set_ID)
  
t.test(
  x = x$`non-retracted`$n, 
  y = x$retracted$n, 
  alternative = "two.sided", 
  paired = F)
cohens_d(x$`non-retracted`$n, x$retracted$n)
```

#### Per Year

Then, as before, a year based comparison.

```{r}
df %>% 
  dplyr::select(Set_ID, Paper_ID, Software_ID, Year) %>%
  group_by(Set_ID, Paper_ID, Year) %>%
  summarize(n=ifelse(is.na(Software_ID), 0, n_distinct(Software_ID)), 
            .groups='drop') %>%
  filter(n > 0) %>%
  distinct() %>%
  group_by(Year, Set_ID) %>%
  summarize(m=mean(n), sd=sd(n), num=n()) %>%
  mutate(CIl=m-(qt(p=.975, df=num-1)*(sd/sqrt(num))), 
         CIu = m+(qt(p=.975, df=num-1)*(sd/sqrt(num)))) %>% 
  ggplot(aes(Year, m)) + 
  geom_line(aes(color=Set_ID)) + 
  geom_ribbon(aes(ymin=CIl,ymax=CIu, fill=Set_ID), alpha=.3) + 
  labs(x='Year', y='Number of Distinct Software', 
    caption = 'Fig. S4: Mean number of distinct software mentioned in articles 
    that contain at least one software, depicted over time for
    retracted and control articles. 95% CIs are indicated by lighter 
    colored areas.') +
  theme(plot.caption = element_text(size=8)) +
    scale_color_manual('Type of Article', 
                      values = c("#2b83ba", "#ff8585")) +
    scale_fill_manual('Type of Article', 
                      values = c("#2b83ba", "#ff8585")) 
```

Similarly, the CIs are quite large due to the reduced sample size, especially for the retracted set and in early years.

#### Per Reason

Again, we view the results per retraction reason. Here, we also combine the two generated plots for a better illustration of the results.

```{r, fig.width=10.5, fig.height=6.5}
df_reason_sampled %>% 
  group_by(set, OriginalReason, Paper_ID) %>%
  drop_na() %>%
  summarize(n=ifelse(is.na(Software_ID), 0, n_distinct(Software_ID))) %>%
  filter(n > 0) %>%
  ungroup() %>%
  distinct() %>%
  group_by(set, OriginalReason) %>%
  summarize(m=mean(n), sd=sd(n), num=n(), min=min(n), 
            max=max(n), median=median(n)) %>% 
  mutate(CIl=m-(qt(p=.975, df=num-1)*(sd/sqrt(num))), 
         CIu = m+(qt(p=.975, df=num-1)*(sd/sqrt(num)))) %>%
  mutate(set=ifelse(set=='non-retracted', "Control", "Retracted")) %>%
  select(OriginalReason, set, m, CIu, CIl) %>% print(., n=16) %>%
  mutate(plot="Number of Distinct Software") %>%
  rename(rel=m) %>%
  mutate(OriginalReason = factor(OriginalReason, levels=reasons)) %>%
  ggplot(aes(OriginalReason, rel)) + 
  geom_point(aes(color=set), position=position_dodge(width=.6)) + 
  geom_errorbar(aes(ymin=CIl, ymax=CIu, color=set), 
                position=position_dodge(width = .6), width=.5) +
  labs(x=element_blank(), y = "Distinct Software")  + 
  scale_color_manual('Type of Article', 
                     values = c("#2b83ba", "#ff8585")) + 
  ylim(1, 3.75) +
  theme(legend.position='none', 
        axis.text.x = element_text(angle=0)) +
  facet_wrap(~ plot, nrow=2, scales='free_y') -> p2

p_out <- p1 / p2 + plot_layout(heights = c(2,1))
ggsave('software_amount.jpg', p_out, width=8, height=5)
p_out  +
  labs(caption = 'Fig S5 (Article Fig. 1.): Software mentions in scholarly articles per 
    retraction reason separated by retracted and corresponding con-
    trol articles. The sets of control papers are constructed by selecting 
    the ten corresponding articles for each retracted article. Top: pro-
    portion of articles that contain at least one software mention. Bottom: 
    average number of software mentions per article with at least one soft-
    ware mention. Error bars indicate 95% CIs.') +
  theme(plot.caption = element_text(size=14))
```

### Software Based Analysis

So far, we have looked at general differences in software usage between retracted and non-retracted articles. Now, we look at differences in usage of specific software. First, how often specific software is used between sets.

#### Overall

We analyze in which percentage of articles individual software is used (within all articles that mention software), and how the distributions vary between sets.

```{r, fig.width=9.5, fig.height=4.5}
df %>%
  filter(Software_Type != "OperatingSystem") %>%
  dplyr::select(Set_ID, Paper_ID, Software_Name, Software_ID) %>%
  mutate(Software_Name=str_replace_all(Software_Name, " - ","-")) %>%
  drop_na() %>%
  distinct() %>%
  group_by(Set_ID) %>%
  mutate(n_articles = n_distinct(Paper_ID)) %>%
  group_by(Set_ID, Software_Name, n_articles, Software_ID) %>%
  count() %>%
  ungroup() %>%
  group_by(Set_ID) %>%
  mutate(rel=n/n_articles) %>%
  mutate(SE = sqrt((rel*(1-rel))/n_articles)) %>%
  mutate(CIl = rel - (1.96*SE), CIu = rel + (1.96*SE)) %>%
  ungroup() %>%
  group_by(Software_ID) %>%
  mutate(s = sum(rel)) %>%
  ungroup() %>%
  slice_max(order_by = s, n = 40) %>%
  mutate(rel=rel*100, CIl=CIl*100, CIu=CIu*100) %>% 
  mutate(Software_Name=reorder(Software_Name, s)) %>%
  select(Set_ID, Software_Name, n, rel, CIl, CIu) %>% print(., n=40) %>%
  mutate(rel=ifelse(Set_ID=='non-retracted', -rel,rel)) %>%
  mutate(CIl=ifelse(Set_ID=='non-retracted', -CIl,CIl)) %>%
  mutate(CIu=ifelse(Set_ID=='non-retracted', -CIu,CIu)) %>%
  mutate(Set_ID=ifelse(Set_ID=='non-retracted', 
                       "Control", 
                       "Retracted")) %>%
  ggplot(aes(rel, Software_Name)) + 
  geom_col(aes(fill=Set_ID)) + 
  geom_errorbar(aes(y=Software_Name, xmin=CIl, xmax=CIu), width=0.8) + 
  geom_text(aes(label=paste0(format(abs(rel),digits=1,nsmall=1), "%"), 
                x = ifelse(abs(CIl)>5, sign(rel)*2.2, CIu + sign(rel)*1.8)), 
            size=3) + 
  labs(x='Relative Number of Articles', 
       y='Disambiguated Software') + 
  scale_fill_manual('Type of Article', 
                    values = c("#2b83ba", "#ff8585")) +
  scale_x_continuous(breaks=c(-.2,-.1,0,.1,.2,.3,.4)*100, 
                     labels=paste0(c(.2,.1,0,.1,.2,.3,.4)*100, "%")) -> 
  p_software

ggsave("software_differences.jpg", p_software, width=8, height = 5)
p_software + 
  labs(caption = 'Fig S6 (Article Fig. 2.): Proportion of retracted and control 
    articles mentioning software out of the top 20 most used 
    software. Error bars indicate 95% CIs.') +
  theme(plot.caption = element_text(size=12))
```

#### Statistics software

We perform the same analyses limited to the most frequently used statistical software because it is the most common software group.

```{r, fig.width=9, fig.height=4.3}
df %>%
  dplyr::select(Set_ID, Paper_ID, Software_Name, Software_ID) %>%
  drop_na() %>%
  distinct() %>%
  group_by(Set_ID) %>%
  mutate(n_articles = n_distinct(Paper_ID)) %>%
  group_by(Set_ID, Software_Name, n_articles, Software_ID) %>%
  count() %>%
  ungroup() %>%
  group_by(Set_ID) %>%
  mutate(rel=n/n_articles) %>%
  mutate(SE = sqrt((rel*(1-rel))/n_articles)) %>%
  mutate(CIl = rel - (1.96*SE), CIu = rel + (1.96*SE)) %>%
  ungroup() %>%
  group_by(Software_ID) %>%
  mutate(s = sum(rel)) %>%
  ungroup() %>%
  inner_join(software_enrichment, by=c(
    'Software_ID'='Software_ID', 
    'Software_Name'='Software_Name')) %>%
  filter(type == "Stat") %>%
  slice_max(order_by = s, n = 30) %>%
  mutate(rel=ifelse(Set_ID=='non-retracted', -rel,rel)) %>%
  mutate(CIl=ifelse(Set_ID=='non-retracted', -CIl,CIl)) %>%
  mutate(CIu=ifelse(Set_ID=='non-retracted', -CIu,CIu)) %>%
  mutate(Set_ID=ifelse(Set_ID=='non-retracted', 
                       "Control", 
                       "Retracted")) %>%
  mutate(Software_Name=reorder(Software_Name, s)) %>%
  mutate(rel=rel*100, CIl=CIl*100, CIu=CIu*100) %>%
  ggplot(aes(rel, Software_Name)) + 
  geom_col(aes(fill=Set_ID)) + 
  geom_errorbar(aes(y=Software_Name, xmin=CIl, xmax=CIu), width=0.8) + 
  geom_text(aes(label=paste0(format(abs(rel),digits=1,nsmall=1), "%"), 
                x = ifelse(abs(CIl)>5, sign(rel)*2.2, CIu + sign(rel)*1.8)), 
            size=3) + 
  labs(x='Relative Number of Articles',
       y='Disambiguated Software',
       caption = 'Fig S7: Proportion of retracted and control 
    articles mentioning software out of the top 15 most used 
    statistical software. Error bars indicate 95% CIs.') +
  scale_fill_manual('Type of Article', 
                    values = c("#2b83ba", "#ff8585")) +
  scale_x_continuous(breaks=c(-.2,-.1,0,.1,.2,.3,.4)*100, 
                     labels=paste0(c(.2,.1,0,.1,.2,.3,.4)*100, "%")) +
  theme(plot.caption = element_text(size=12))
```

#### Per Reason

Next, we look at individual software split per retraction reason.

```{r, fig.width=10, fig.height=7.5}
df_reason_sampled %>%
  select(set, Paper_ID, Software_Name, Software_ID, OriginalReason) %>%
  drop_na() %>%
  distinct() %>%
  group_by(set, OriginalReason) %>%
  mutate(n_articles = n_distinct(Paper_ID)) %>%
  group_by(set, OriginalReason, Software_Name, n_articles, Software_ID) %>%
  count() %>%
  group_by(set, OriginalReason) %>%
  mutate(rel=n/n_articles) %>%
  mutate(SE = sqrt((rel*(1-rel))/n_articles)) %>%
  mutate(CIl = rel - (1.96*SE), CIu = rel + (1.96*SE)) %>%
  ungroup() %>%
  group_by(Software_ID, OriginalReason) %>%
  mutate(s = sum(rel)) %>%
  ungroup() %>%
  group_by(OriginalReason) %>%
  slice_max(order_by = s, n = 20) %>%
  ungroup() %>%
  mutate(rel=ifelse(set=='non-retracted', -rel,rel)) %>%
  mutate(CIl=ifelse(set=='non-retracted', -CIl,CIl)) %>%
  mutate(CIu=ifelse(set=='non-retracted', -CIu,CIu)) %>%
  mutate(Software_Name=reorder(Software_Name, s)) %>% 
  mutate(rel=rel*100, CIl=CIl*100, CIu=CIu*100) %>%
  mutate(OriginalReason = factor(OriginalReason, levels=reasons)) -> 
  tmp_df

tmp_df %>%
  filter(Software_Name == 'TargetScan', OriginalReason == 'PaperMill') %>%
  select(set, OriginalReason, Software_Name, rel, CIl, CIu) %>%
  mutate(across(where(is.numeric), round, 1))%>%
  mutate(across(where(is.numeric), abs))

tmp_df %>%
  filter(Software_Name == 'SPSS', OriginalReason == 'PaperMill') %>%
  select(set, OriginalReason, Software_Name, rel, CIl, CIu) %>%
  mutate(across(where(is.numeric), round, 1))%>%
  mutate(across(where(is.numeric), abs))

tmp_df %>%
  ggplot(aes(rel,Software_Name)) + 
  geom_col(aes(fill=set)) + 
  geom_errorbar(aes(y=Software_Name, xmin=CIl, xmax=CIu)) + 
  labs(x='Relative Number of Mentions', 
       y='Disambiguated Software',
       caption = 'Fig S8: Proportion of retracted and control articles mentioning 
    software out of the top 10 most used software per retraction 
    reason. A separate control set is constructed for each retrac-
    tion reasons by selecting the ten corresponding articles for 
    each retracted paper.Error bars indicate 95% CIs.') + 
  scale_fill_manual('Type of Article', 
                    values = c("#2b83ba", "#ff8585")) +
  theme(legend.position = 'top',
        plot.caption = element_text(size=14)) +
  scale_x_continuous(breaks=c(-.2,0,.2,.4,.6,.8)*100, 
                     labels=paste0(c(.2,0,.2,.4,.6,.8)*100, "%")) +
  facet_wrap(scales="free_y", ~ OriginalReason)
```

### Software Distribution

We look at how software is distributed within articles by analyzing in what proportion of articles any of the top n software appears. This gives us an estimate of how diverse the used software is.

```{r, fig.width=8}
get_nums <- function(df, num) {
  df %>%
    select(Set_ID, Paper_ID, Software_ID, Software_Name) %>%
    drop_na(Software_ID) %>%
    filter(Set_ID == 'non-retracted') %>%
    distinct() %>%
    group_by(Software_ID, Software_Name) %>%
    summarize(n=n(), .groups = 'drop') %>%
    arrange(desc(n)) %>%
    slice_head(n = num) -> top_n_control
  
  df %>%
    select(Set_ID, Paper_ID, Software_ID, Software_Name) %>%
    drop_na(Software_ID) %>%
    filter(Set_ID == 'retracted') %>%
    distinct() %>%
    group_by(Software_ID, Software_Name) %>%
    summarize(n=n(), .groups = 'drop') %>%
    arrange(desc(n)) %>%
    slice_head(n = num) -> top_n_retracted
  
  df %>%
    select(Set_ID, Paper_ID, Software_ID, Software_Name) %>%
    drop_na(Software_ID) %>%
    filter(Set_ID == 'non-retracted') %>%
    group_by(Set_ID) %>%
    mutate(o=n_distinct(Paper_ID)) %>%
    distinct() %>%
    filter(Software_ID %in% top_n_control$Software_ID) %>%
    group_by(Set_ID, o) %>%
    summarize(n = n_distinct(Paper_ID), .groups = 'drop_last') %>%
    mutate(rel = n / o) -> res1
  
  df %>%
    select(Set_ID, Paper_ID, Software_ID, Software_Name) %>%
    drop_na(Software_ID) %>%
    filter(Set_ID == 'retracted') %>%
    group_by(Set_ID) %>%
    mutate(o=n_distinct(Paper_ID)) %>%
    distinct() %>%
    filter(Software_ID %in% top_n_retracted$Software_ID) %>%
    group_by(Set_ID, o) %>%
    summarize(n = n_distinct(Paper_ID), .groups = 'drop_last') %>%
    mutate(rel = n / o) -> res2
  
  rbind(res1, res2) %>%
    mutate(num_id=num)
}

lapply(1:76, function(i){get_nums(df, i)}) %>% bind_rows() -> out_df

out_df %>%
  ggplot(aes(x=num_id, y=rel, group=Set_ID, color=Set_ID)) +
  geom_line() +
  geom_point() + 
  labs(x='Top n software', 
       y='Amount of Articles', 
       caption = 'Fig S9: Relative amount of articles mentioning at least one of 
       the top n software out of all articles that mention software.') + 
  scale_y_continuous(limits = c(0, 0.90), 
                     breaks=c(0, 0.25, 0.50, 0.75)) +
  scale_color_manual('Type of Article', 
                     values = c("#2b83ba", "#ff8585")) +
  theme(plot.caption = element_text(size=11))

# Getting breakpoints
out_df %>%
  filter(rel > 0.25) %>%
  group_by(Set_ID) %>%
  slice_min(order_by=rel, n=1) %>%
  mutate(value = .25)

out_df %>%
  filter(rel > 0.5) %>%
  group_by(Set_ID) %>%
  slice_min(order_by=rel, n=1) %>%
  mutate(value = .5)

out_df %>%
  filter(rel > 0.75) %>%
  group_by(Set_ID) %>%
  slice_min(order_by=rel, n=1) %>%
  mutate(value = .75)
```

### Software Names and Spelling Variations

Different spelling variations and names are used to refer to the same software. We analyze if there is a trend towards using the most common software name.

```{r, fig.width=8.5}
df %>%
  filter(Set_ID == 'retracted') %>%
  select(Set_ID, Software_ID, Software_Name, Software_String) %>%
  drop_na(Software_ID) %>%
  group_by(Set_ID, Software_ID, Software_Name) %>% 
  summarize(n = n_distinct(Software_String), .groups = 'drop') %>%
  filter(n > 8) -> software_to_compare

df %>%
  drop_na(Software_ID) %>%
  filter(Software_ID %in% software_to_compare$Software_ID) %>%
  select(Set_ID, Software_ID, Software_Name, Software_String) %>%
  group_by(Set_ID, Software_ID, Software_Name, Software_String) %>% 
  summarize(n = n(), .groups = 'drop_last') %>%
  arrange(desc(n)) %>%
  mutate(rel = n/sum(n)) %>%
  mutate(n = sum(n)) %>%
  mutate(SE = sqrt((rel*(1-rel))/n)) %>%
  mutate(CIl = rel - (1.96*SE), CIu = rel + (1.96*SE)) %>%
  mutate(rel = rel*100, CIu=CIu*100, CIl=CIl*100) %>%
  slice_head(n=1) %>%
  ungroup() %>%
  ggplot(aes(x=Software_Name, y=rel, group=Set_ID, color=Set_ID)) +
  geom_errorbar(aes(ymin=CIl, ymax=CIu), 
                position=position_dodge(width = .6), width=.5) +
  geom_point(position=position_dodge(width=.6)) +
  scale_color_manual('Type of Article', 
                     values = c("#2b83ba", "#ff8585")) +
  labs(x='Disambiguated Software', 
       y='Amount of Mentions', 
       caption = 'Fig S10: Relative amount of articles mentioning a specific software
    and referring to it by its most commonly used name.') +
  theme(plot.caption = element_text(size=11))
```

Taking a closer look at the names used for SPSS and ImageJ, which we found to be differently mentioned between retracted and control set.

```{r}
df %>%
  filter(Software_Name %in% c('SPSS', 'ImageJ')) %>%
  select(Set_ID, Paper_ID, Software_Name, Software_String) %>%
  distinct() %>%
  group_by(Set_ID, Software_Name, Software_String) %>%
  summarize(n = n(), .groups = 'drop_last') %>%
  mutate(rel = n / sum(n)) %>%
  mutate(n = sum(n)) %>%
  mutate(SEM=sqrt((rel * (1-rel))/n), 
         MoE = sqrt((rel * (1-rel))/n) * 1.96) %>%
  mutate(CIu = rel + MoE, CIl = rel-MoE)  %>%
  mutate(rel = rel*100, CIu=CIu*100, CIl=CIl*100) %>%
  select(Set_ID, Software_Name, Software_String, rel, CIl, CIu) %>%
  slice_max(order_by = rel, n = 5) %>%
  arrange(desc(rel), .by_group=TRUE) %>% print(., n=21)
```

### Free and Open Source Software

We analyze the use of free vs commercial software and the use of open- vs closed-source software between retracted and control set.

#### Overall

First, we perform an overall compare between sets.

##### Free

```{r}
df %>%
  dplyr::select(Set_ID, Paper_ID, Software_ID, Software_Name) %>%
  distinct() %>%
  inner_join(software_enrichment, by=c(
    'Software_ID'='Software_ID', 
    'Software_Name'='Software_Name')) %>%
  group_by(Set_ID, free) %>%
  summarize(n=n(), .groups = 'drop_last') %>%
  mutate(rel = n / sum(n)) %>%
  mutate(n = sum(n)) %>%
  ungroup() %>%
  filter(free==1) %>%
  mutate(SEM=sqrt((rel * (1-rel))/n), 
         MoE = sqrt((rel * (1-rel))/n) * 1.96) %>%
  mutate(CIu = rel + MoE, CIl = rel-MoE) %>%
  select(Set_ID, free, n, rel, CIl, CIu)
```

##### Open Source

```{r}
df %>%
  dplyr::select(Set_ID, Paper_ID, Software_ID, Software_Name) %>%
  distinct() %>%
  inner_join(software_enrichment, by=c(
    'Software_ID'='Software_ID', 
    'Software_Name'='Software_Name')) %>%
  group_by(Set_ID, source) %>%
  summarize(n=n(), .groups = 'drop_last') %>%
  mutate(rel = n / sum(n)) %>%
  ungroup() %>%
  filter(source==1) %>%
  mutate(SEM=sqrt((rel * (1-rel))/n), 
         MoE = sqrt((rel * (1-rel))/n) * 1.96) %>%
  mutate(CIu = rel + MoE, CIl = rel-MoE) %>%
  select(Set_ID, source, n, rel, CIl, CIu)
```

```{r, fig.height=3.5}
df %>%
  dplyr::select(Set_ID, Paper_ID, Software_ID, Software_Name) %>%
  distinct() %>%
  inner_join(software_enrichment, by=c(
    'Software_ID'='Software_ID', 
    'Software_Name'='Software_Name')) %>%
  mutate(Set_ID=ifelse(Set_ID=='retracted', 
                       "Retracted", 
                       "Control")) %>%
  mutate(Set_ID=factor(Set_ID, levels= c("Control","Retracted"))) %>%
  mutate(free=ifelse(free, 'Free', 'Commercial')) %>%
  mutate(source=ifelse(source, 'Open Source', 'Closed Source')) %>%
  pivot_longer(c('free', 'source'), names_to="open_source") %>%
  mutate(open_source= ifelse('free'==open_source, 
                             "Software availability", 
                             "Source availablity")) %>%
  group_by(Set_ID, open_source, value) %>%
  mutate(value=factor(value, levels=c('Free', 
                                      'Commercial', 
                                      'Open Source', 
                                      'Closed Source')))%>%
  summarize(n=n(), .groups = 'drop_last') %>%
  mutate(rel = n / sum(n)) %>%
  ungroup() %>%
  group_by(Set_ID, open_source) %>%
  mutate(n = sum(n)) %>%
  mutate(SEM=sqrt((rel * (1-rel))/n), 
         MoE = sqrt((rel * (1-rel))/n) * 1.96) %>%
  mutate(CIu = rel + MoE, CIl = rel-MoE) %>%
  mutate(rel = rel*100, CIu=CIu*100, CIl=CIl*100) %>%
  ggplot(aes(x=value, rel)) + 
  geom_point(aes(color=Set_ID), 
             position = position_dodge(width = 0.7)) +
  geom_errorbar(aes(ymin=CIl, ymax=CIu, color=Set_ID), 
                position = position_dodge(width = 0.7), width=0.6) + 
  scale_y_continuous(limits=c(0,75), breaks=c(0,.2,.4,.6)*100, 
                     labels=paste0(c(0,.2,.4,.6)*100, "%")) +
  facet_grid(~ open_source, scales='free_x') +
  scale_color_manual("Type of Article", 
                     values = c("#2b83ba", "#ff8585")) +
  labs(y="Proportion of Software") + 
  theme(axis.title.x = element_blank()) ->
  p_source

ggsave("Open_source_software.jpg", p_source, width=4.3, height = 3)
p_source +
  labs(caption = 'Fig S11 (Article Fig. 3.): Proportion of free or open source 
    software across retracted and control articles. Error bars 
    indicate 95% CIs.') +
  theme(plot.caption = element_text(size=8))
```

###### Statistical Test

We further perform a statistical test to investigate if there is a difference in free and open source software usage between retracted and control articles. In this context, we observed that there is a relation between the amount of free (and open source) software and the number of software used within an article, where the ratio of free (and open source) software increases with the number of software per article. Therefore, we include this number of software in an article as a covariate in tests on free (and open source) software.

```{r}
df %>%
  dplyr::select(Set_ID, Paper_ID, Software_ID, 
                Software_Name, Control_Sample_Origin) %>%
  distinct() %>%
  inner_join(software_enrichment, by=c(
    'Software_ID'='Software_ID', 
    'Software_Name'='Software_Name')) %>%
  group_by(Paper_ID, Control_Sample_Origin) %>%
  summarize(
    n_free = sum(free) / n(), 
    n_source = sum(source) / n(), 
    n=n(), 
    .groups = 'drop_last') -> df_t

df_t %>%
  group_by(n) %>%
  summarize(
    m_free = mean(n_free), 
    sd_free=sd(n_free),
    m_source = mean(n_source), 
    sd_source=sd(n_source)
    ) %>%
  pivot_longer(-n, 
               names_to=c('value_type', 'availability'), 
               names_sep='_') %>%
  pivot_wider(names_from = value_type, values_from = value) %>%
  ggplot(aes(n, m, color=availability)) + 
  geom_point() +
  geom_line() +
  geom_ribbon(aes(ymin=pmax(0, m-sd), ymax=pmin(1, m+sd)), alpha=.1)
```

We use a GLM and include the retraction state and the number of software in the article as covariates to predict the availability of a software, and further include their interactions for completeness. Effect sizes are then estimated through odds ratios.

```{r}
df %>%
  dplyr::select(Set_ID, Paper_ID, Software_ID, 
                Software_Name, Control_Sample_Origin) %>%
  distinct() %>%
  inner_join(software_enrichment, by=c(
    'Software_ID'='Software_ID', 
    'Software_Name'='Software_Name')) %>%
  group_by(Set_ID, Paper_ID, Control_Sample_Origin) %>%
  mutate(n = n()) %>%
  ungroup() ->  df_tt

model <- glm(
  formula=free~Set_ID+n+Set_ID*n, 
  data=mutate(df_tt, Set_ID=factor(Set_ID, c('retracted', 'non-retracted'))), 
  family = binomial(link="logit"))
summary(model)
exp(summary(model)$coefficients["n",1] + qnorm(c(0.025,0.5,0.975)) * 
      summary(model)$coefficients["n",2])
exp(summary(model)$coefficients["Set_IDnon-retracted",1] + qnorm(c(0.025,0.5,0.975)) * 
      summary(model)$coefficients["Set_IDnon-retracted",2])

model <- glm(
  formula=source~Set_ID+n+Set_ID*n, 
  data=mutate(df_tt, Set_ID=factor(Set_ID, c('retracted', 'non-retracted'))), 
  family = binomial(link="logit"))
summary(model)
exp(summary(model)$coefficients["n",1] + qnorm(c(0.025,0.5,0.975)) * 
      summary(model)$coefficients["n",2])
exp(summary(model)$coefficients["Set_IDnon-retracted",1] + qnorm(c(0.025,0.5,0.975)) * 
      summary(model)$coefficients["Set_IDnon-retracted",2])
```

#### Per Reason

Then, we extend the analyses concerning individual retraction reasons.

```{r, fig.height=15}
df_reason_sampled %>%
  select(set, Paper_ID, Software_ID, Software_Name, OriginalReason) %>%
  distinct() %>%
  inner_join(software_enrichment, by=c(
    'Software_ID'='Software_ID', 
    'Software_Name'='Software_Name')) %>%
  mutate(Set_ID=ifelse(set=='retracted', 
                       "Retracted", 
                       "Control")) %>%
  mutate(Set_ID=factor(set, levels= c("Control","Retracted"))) %>%
  mutate(free=ifelse(free, 'Free', 'Commercial')) %>%
  mutate(source=ifelse(source, 
                       'Open Source', 
                       'Closed Source')) %>%
  pivot_longer(c('free', 'source'), names_to="open_source") %>%
  mutate(open_source= ifelse('free'==open_source, 
                             "Software availability", 
                             "Source availablity")) %>%
  group_by(set, OriginalReason, open_source, value) %>%
  mutate(value=factor(value, levels=c('Free', 
                                      'Commercial', 
                                      'Open Source', 
                                      'Closed Source')))%>%
  summarize(n=n(), .groups = 'drop_last') %>%
  mutate(rel = n / sum(n)) %>%
  mutate(n = sum(n)) %>%
  ungroup() %>%
  mutate(SEM=sqrt((rel * (1-rel))/n), 
         MoE = sqrt((rel * (1-rel))/n) * 1.96) %>%
  mutate(CIu = rel + MoE, CIl = rel-MoE) %>%
  mutate(rel = rel*100, CIu=CIu*100, CIl=CIl*100) %>%
  select(set, OriginalReason, open_source, value, n, rel, CIl, CIu) %>% 
  print(., n=64) %>%
  ggplot(aes(x=value, rel)) + 
  geom_point(aes(color=set), 
             position = position_dodge(width = 0.7)) +
  geom_errorbar(aes(ymin=CIl, ymax=CIu, color=set), 
                position = position_dodge(width = 0.7), width=0.6) +
  facet_grid(OriginalReason ~ open_source , scales='free_x') +
  scale_color_manual("Type of Article", 
                     values=c("#2b83ba", "#ff8585")) +
  labs(y="Proportion of Software",
       caption='Fig S12: Proportion of free or open source software across
    retracted and control articles per retraction reason. A sep-
    arate control set is constructed for each retraction reasons 
    by selecting the ten corresponding articles for each retrac-
    ted paper. Error bars indicate 95% CIs.') + 
  theme(axis.title.x = element_blank(),
        plot.caption = element_text(size=14)) 

```

### Software Type

We analyze the difference in software type usage between retracted and control articles based on the types: Application, PlugIn, Programming Environment, and Operating System.

```{r, fig.width=8}
df %>%
  dplyr::select(Set_ID, Paper_ID, Software_ID, Software_Type) %>%
  drop_na() %>%
  distinct() %>%
  group_by(Set_ID, Software_Type) %>%
  summarize(n=n(), .groups = 'drop_last') %>%
  group_by(Set_ID) %>%
  mutate(rel=n/sum(n)) %>%
  group_by(Set_ID) %>%
  mutate(n=sum(n)) %>%
  ungroup() %>%
  mutate(SEM=sqrt((rel * (1-rel))/n),
         MoE = sqrt((rel * (1-rel))/n) * 1.96) %>%
  mutate(CIu = rel + MoE, CIl = rel-MoE) %>%
  mutate(rel=rel*100, CIl=CIl*100, CIu=CIu*100) %>%
  mutate(Set_ID=ifelse(Set_ID=='non-retracted', 
                       "Control", 
                       "Retracted")) %>%
  select(Set_ID, Software_Type, n, rel, CIl, CIu) %>% 
  print(., n=8) %>%
  ggplot(aes(rel, Software_Type)) +
  geom_col(aes(fill=Set_ID), position='dodge') +
  geom_errorbar(aes(y=Software_Type, xmin=CIl, xmax=CIu, group=Set_ID), 
                position='dodge') +
  geom_text(aes(x = ifelse(abs(CIl)>10, CIl - 4, CIu + 3.5), 
                y=Software_Type, 
                group=Set_ID, 
                label=paste0(format(abs(rel),digits=1,nsmall=1), "%")), 
            position = position_dodge(width = .9), size=3) + 
  scale_fill_manual('Type of Article', 
                    values = c("#2b83ba", "#ff8585")) +
  theme(legend.position = 'top',
        plot.caption = element_text(size=10)) +
  labs(x='Proportion of Mentions',
       y='Software Type',
       caption = 'Fig S13: Proportion of software types on overall software mentions
    between retracted and control articles.')
```

## Citation Quality

After exploring the software landscape, we analyze the citation quality for software.

### Overall

First, we directly compare the sets.

```{r, fig.height=3.5}
names <- c("No Info", "Incomplete Info", 
           "Informal Citation", "Formal Citation")

df %>%
  select(Set_ID, Paper_ID, Software_ID, Version, Developer, Citation) %>%
  filter(!is.na(Software_ID)) %>%
  group_by(Set_ID, Paper_ID, Software_ID) %>%
  summarize(Version=any(Version), 
            Developer=any(Developer), 
            Citation=any(Citation), 
            .groups = 'drop_last') %>%
  mutate(version_and_developer = Version & Developer & ! Citation, 
         no_citation_info = ! Version & ! Developer & ! Citation,
         version_or_developer = (Version & ! Developer & ! Citation) | 
           (Developer & ! Version & ! Citation),
         citation_p = Citation) %>%
  dplyr::select(-Version, -Developer, -Citation) %>%
  rename(Citation=citation_p) %>%
  distinct() %>%
  group_by(Set_ID) %>%
  summarize(`No Info` = sum(no_citation_info),
            `Incomplete Info` = sum(version_or_developer),
            `Formal Citation` = sum(Citation),
            `Informal Citation` = sum(version_and_developer),
            n = n(),
            .groups = 'drop_last') %>%
  pivot_longer(c(`No Info`, `Incomplete Info`, 
                 `Formal Citation`, `Informal Citation`)) %>%
  mutate(rel = value/n) %>%
  mutate(SEM=sqrt((rel * (1-rel))/n), 
         MoE = sqrt((rel * (1-rel))/n) * 1.96) %>%
  mutate(CIl = rel-MoE, CIu = rel + MoE) %>% 
  group_by(name) %>%
  mutate(order=sum(rel))  %>%
  ungroup() %>%
  mutate(name=factor(name, levels=names)) %>%
  mutate(rel=rel*100, CIu=CIu*100, CIl=CIl*100) %>%
  mutate(Set_ID = ifelse('non-retracted'==Set_ID, 
                         "Control", 
                         "Retracted")) %>%
  select(Set_ID, name, value, rel, CIl, CIu) %>% 
  print(.) %>%
  ggplot(aes(name, rel)) + 
  geom_point(aes(x=name, color=Set_ID), 
             position=position_dodge(width=.7)) + 
  geom_errorbar(aes(x=name, ymin=CIl, ymax=CIu, color=Set_ID), 
                position=position_dodge(width=.7), width=.5) + 
  labs(x='Relative Number of Mentions', 
       y='Relative Number of Mentions',
       caption = 'Fig S14: Proportion of software across different levels of citation
    completeness, separated by retracted and control articles. No Info: 
    Neither the version, nor the developer of a software are provided; 
    Incomplete Info: Either version or developer is provided; Informal 
    Citation: Version and developer are provided; Formal citation: soft-
    ware mention is accompanied by bibliographic citation. Error bars 
    indicate 95% CIs.') + 
  scale_color_manual('Type of Article', 
                     values = c("#2b83ba", "#ff8585")) +
  scale_y_continuous(limits=c(0,45), 
                     breaks=c(0, 20, 40), 
                     labels=c("0%", "20%", "40%")) + 
  theme(legend.position = 'top', 
        axis.title.x = element_blank(),
        plot.margin = unit(c(0,1,0,1), 'mm'),
        plot.caption = element_text(size=8))

```

### Free and Commercial Software

Then, we look at the citation quality divided by free and commercial software because prior work has shown that there are differences in their citation [@Du2022].

```{r, fig.height=4, fig.width=8}

df %>%
  select(Set_ID, Paper_ID, Software_ID, Software_Name, 
         Version, Developer, Citation) %>%
  filter(!is.na(Software_ID)) %>%
  inner_join(software_enrichment, by=c(
    'Software_ID'='Software_ID', 
    'Software_Name'='Software_Name')) %>%
  group_by(Set_ID, free, Paper_ID, Software_ID) %>%
  summarize(Version=any(Version), 
            Developer=any(Developer), 
            Citation=any(Citation), 
            .groups = 'drop_last') %>%
  mutate(version_and_developer = Version & Developer & ! Citation, 
         no_citation_info = ! Version & ! Developer & ! Citation,
         version_or_developer = (Version & ! Developer & ! Citation) | 
           (Developer & ! Version & ! Citation),
         citation_p = Citation) %>%
  select(-Version, -Developer, -Citation) %>%
  rename(Citation=citation_p) %>%
  distinct() %>%
  group_by(Set_ID, free) %>%
  summarize(`No Info` = sum(no_citation_info),
            `Incomplete Info` = sum(version_or_developer),
            `Formal Citation` = sum(Citation),
            `Informal Citation` = sum(version_and_developer),
            n = n(),
            .groups = 'drop_last') %>%
  pivot_longer(c(`No Info`, `Incomplete Info`, 
                 `Formal Citation`, `Informal Citation`)) %>%
  mutate(rel = value/n) %>%
  mutate(SEM=sqrt((rel * (1-rel))/n), 
         MoE = sqrt((rel * (1-rel))/n) * 1.96) %>%
  mutate(CIl = rel-MoE, CIu = rel + MoE) %>%
  group_by(name) %>%
  mutate(order=sum(rel))  %>%
  ungroup() %>%
  mutate(name=factor(name, levels=names)) %>%
  mutate(free=ifelse(free, "Free", "Commercial")) %>%
  mutate(rel=rel*100, CIu=CIu*100, CIl=CIl*100) %>%
  mutate(Set_ID = ifelse('non-retracted'==Set_ID, 
                         "Control", 
                         "Retracted")) %>%
  select(Set_ID, free, name, value, rel, CIl, CIu) %>% 
  print(.) ->
  df_ccomplete

df_ccomplete %>%
  ggplot(aes(free,rel)) + 
  geom_point(aes(x=free, color=Set_ID), 
             position=position_dodge(width=.7)) + 
  geom_errorbar(aes(x=free, ymin=CIl, ymax=CIu, color=Set_ID), 
                position=position_dodge(width=.7), width=.5) + 
  labs(x='Relative Number of Mentions', 
       y='Relative Number of Mentions') + 
  scale_color_manual('Type of Article', 
                     values = c("#2b83ba", "#ff8585")) +
  scale_y_continuous(breaks=c(0, 20, 40), 
                     labels=c("0%", "20%", "40%")) + 
  theme(legend.position = 'top', 
        axis.title.x = element_blank(),
        plot.margin = unit(c(0,1,0,1), 'mm')) +
  facet_grid(~name) -> pp

ggsave("one_column_plot.jpg", pp, width = 8, height = 3)
pp +
  labs(caption = 'Fig S15 (Article Fig. 4): Proportion of software across different
    levels of citation completeness, separated by retracted and control
    articles and between free and commercial software. No Info: Neither 
    the version, nor the developer of a software are provided; Incomplete
    Info: Either version or developer is provided; Informal Citation: 
    Version and developer are provided; Formal citation: software mention 
    is accompanied by bibliographic citation. Error bars indicate 95% CIs.') +
  theme(plot.caption = element_text(size=10))
```

### Open and Closed Source Software

We perform the same analysis for open and closed source software expecting similar results as the attributes free and open source are strongly correlated.

```{r, fig.height=4, fig.width=8}
df %>%
  dplyr::select(Set_ID, Paper_ID, Software_ID, Software_Name, 
                Version, Developer, Citation) %>%
  filter(!is.na(Software_ID)) %>%
  inner_join(software_enrichment, by=c(
    'Software_ID'='Software_ID', 
    'Software_Name'='Software_Name')) %>%
  group_by(Set_ID, source, Paper_ID, Software_ID) %>%
  summarize(Version=any(Version), 
            Developer=any(Developer), 
            Citation=any(Citation), 
            .groups = 'drop_last') %>%
  mutate(version_and_developer = Version & Developer & ! Citation, 
         no_citation_info = ! Version & ! Developer & ! Citation,
         version_or_developer = (Version & ! Developer & ! Citation) | 
           (Developer & ! Version & ! Citation),
         citation_p = Citation) %>% 
  dplyr::select(-Version, -Developer, -Citation) %>%
  rename(Citation=citation_p) %>%
  distinct() %>%
  group_by(Set_ID, source) %>%
  summarize(`No Info` = sum(no_citation_info),
            `Incomplete Info` = sum(version_or_developer),
            `Formal Citation` = sum(Citation),
            `Informal Citation` = sum(version_and_developer),
            n = n(),
            .groups = 'drop_last') %>%
  pivot_longer(c(`No Info`, `Incomplete Info`, 
                 `Formal Citation`, `Informal Citation`)) %>%
  mutate(rel = value/n) %>%
  mutate(SEM=sqrt((rel * (1-rel))/n), 
         MoE = sqrt((rel * (1-rel))/n) * 1.96) %>%
  mutate(CIl = rel-MoE, CIu = rel + MoE) %>%
  group_by(name) %>%
  mutate(order=sum(rel))  %>%
  ungroup() %>%
  mutate(name=factor(name, levels=names)) %>%
  mutate(source=ifelse(source, "Open", "Closed")) %>%
  mutate(rel=rel*100, CIu=CIu*100, CIl=CIl*100) %>%
  mutate(Set_ID = ifelse('non-retracted'==Set_ID, 
                         "Control", 
                         "Retracted")) %>%
  select(Set_ID, source, name, value, rel, CIl, CIu) %>% 
  print(.) ->
  df_ccomplete

df_ccomplete %>%
  ggplot(aes(source,rel)) + 
  geom_point(aes(x=source, color=Set_ID), 
             position=position_dodge(width=.7)) + 
  geom_errorbar(aes(x=source, ymin=CIl, ymax=CIu, color=Set_ID), 
                position=position_dodge(width=.7), width=.5) + 
  labs(x='Relative Number of Mentions', 
       y='Relative Number of Mentions', 
       caption = 'Fig S16: Proportion of software across different levels of citation
    completeness, separated by retracted and control articles and between 
    open-source and closed-source software. No Info: Neither the version, 
    nor the developer of a software are provided; Incomplete Info: Either 
    version or developer is provided; Informal Citation: Version and devel-
    oper are provided; Formal citation: software mention is accompanied 
    by bibliographic citation. Error bars indicate 95% CIs.') + 
  scale_color_manual('Type of Article', 
                     values = c("#2b83ba", "#ff8585")) +
  scale_y_continuous(breaks=c(0, 20, 40), 
                     labels=c("0%", "20%", "40%")) + 
  theme(legend.position = 'top', 
        axis.title.x = element_blank(),
        plot.margin = unit(c(0,1,0,1), 'mm'),
        plot.caption = element_text(size=11)) +
  facet_grid(~name) 
```

###### Statistical Test

We include a further statistical test to test whether the citation quality in terms of formal software citation differs between retracted and control articles. We use a GLM and include retraction state, number of software, and availability (free vs commercial or open source vs close source) as covariates to predict whether a software was formally cited. We further include their interaction, particularly, retraction state and availability, which we know to be related from a prior test. Finally, we estimate effect sizes through odds ratios.

```{r}
df %>%
  select(Set_ID, Paper_ID, Software_ID, Software_Name, Version, 
         Developer, Citation, Control_Sample_Origin) %>%
  filter(!is.na(Software_ID)) %>%
  inner_join(software_enrichment, by=c(
    'Software_ID'='Software_ID', 
    'Software_Name'='Software_Name')) %>%
  group_by(Set_ID, Paper_ID,  Software_ID, 
           Control_Sample_Origin, free, source) %>%
  summarise(citation=any(Citation), .groups = 'drop') %>%
  group_by(Set_ID, Paper_ID, Control_Sample_Origin) %>%
  mutate(n=n()) %>% 
  ungroup() -> df_tt
  
model <- glm(
  citation~Set_ID+source+free+n+.*.,
  data=select(
    mutate(
      df_tt, 
      Set_ID=factor(Set_ID, c('retracted', 'non-retracted'))), 
    -c(Paper_ID, Software_ID, Control_Sample_Origin)), 
  family=binomial(link='logit'))
summary(model)
exp(summary(model)$coefficients["n",1] + qnorm(c(0.025,0.5,0.975)) * 
      summary(model)$coefficients["n",2])
exp(summary(model)$coefficients["free",1] + qnorm(c(0.025,0.5,0.975)) * 
      summary(model)$coefficients["free",2])
exp(summary(model)$coefficients["source",1] + qnorm(c(0.025,0.5,0.975)) * 
      summary(model)$coefficients["source",2])
exp(summary(model)$coefficients["Set_IDnon-retracted",1] + qnorm(c(0.025,0.5,0.975)) * 
      summary(model)$coefficients["Set_IDnon-retracted",2])
exp(summary(model)$coefficients["source:free",1] + qnorm(c(0.025,0.5,0.975)) * 
      summary(model)$coefficients["source:free",2])
```

### Per Reason (Free and Commercial)

Then, we extend the analysis to cover retraction reasons.

```{r, fig.width=11, fig.height=8.5}
df_reason_sampled %>%
  filter(!is.na(Software_ID)) %>%
  inner_join(software_enrichment, by=c(
    'Software_ID'='Software_ID', 
    'Software_Name'='Software_Name')) %>%
  group_by(set, free, OriginalReason, Paper_ID, Software_ID) %>%
  summarize(Version=any(Version), 
            Developer=any(Developer), 
            Citation=any(Citation), 
            .groups = 'drop_last') %>%
  distinct() %>%
  mutate(version_and_developer = Version & Developer & ! Citation, 
         no_citation_info = ! Version & ! Developer & ! Citation,
         version_or_developer = (Version & ! Developer & ! Citation) | 
           (Developer & ! Version & ! Citation),
         citation_p = Citation) %>% 
  dplyr::select(-Version, -Developer, -Citation) %>%
  rename(Citation=citation_p) %>%
  distinct() %>%
  group_by(set, free, OriginalReason) %>%
  summarize(`No Info` = sum(no_citation_info),
            `Incomplete Info` = sum(version_or_developer),
            `Formal Citation` = sum(Citation),
            `Informal Citation` = sum(version_and_developer),
            n = n(),
            .groups = 'drop_last') %>%
  pivot_longer(c(`No Info`, `Incomplete Info`, 
                 `Formal Citation`, `Informal Citation`)) %>%
  mutate(rel = value/n) %>%
  mutate(SEM=sqrt((rel * (1-rel))/n), 
         MoE = sqrt((rel * (1-rel))/n) * 1.96) %>%
  mutate(CIl = rel-MoE, CIu = rel + MoE) %>%
  group_by(name) %>%
  ungroup() %>%
  mutate(name=factor(name, levels=names)) %>%
  mutate(rel=rel*100, CIu=CIu*100, CIl=CIl*100) %>%
  mutate(free=ifelse(free, "Free", "Commercial")) %>%
  mutate(set = ifelse('non-retracted'==set, 
                      "Control", 
                      "Retracted")) %>%
  mutate(OriginalReason = factor(OriginalReason, levels=reasons)) %>%
  select(set, free, OriginalReason, name, rel, CIl, CIu) %>% 
  print(., n=128) ->
  df_creasons 

filter(df_creasons, OriginalReason %in% c("Error", 
                                          "Investigation", 
                                          "Plagiarism", 
                                          "SelfPlagiarism")) %>%
  ggplot(aes(free,rel)) + 
  geom_point(aes(color=set), 
             position=position_dodge(width=.7)) + 
  geom_errorbar(aes(x=free, ymin=CIl, ymax=CIu, color=set), 
                position=position_dodge(width=.7), width=.5) + 
  labs(x='Relative Number of Mentions', 
       y='Relative Number of Mentions') + 
  scale_color_manual('Type of Article', 
                     values = c("#2b83ba", "#ff8585")) +
  theme(legend.position = 'top', 
        axis.title.x = element_blank(), axis.text.y = element_blank(), 
        axis.ticks.y.left = element_blank(),
        axis.text.y.right = element_text(hjust=.55),
        axis.title.y.right = element_blank(),
        plot.margin = unit(c(0,1,0,0), 'mm'),
        )  +
  facet_grid(OriginalReason~name, switch='y') +
    scale_y_continuous(
      limits = c(-2, 82), 
      breaks=c(0, 20, 40, 60, 80), 
      labels=c("0%", "20%", "40%", "60%", "80%"), 
      sec.axis = sec_axis(~., breaks=c(0, 20, 40, 60, 80), 
                          labels=c("0%", "20%", "40%", "60%", "80%"))) ->
    p_creasons1

filter(df_creasons, OriginalReason %in% c("Misconduct", 
                                          "PaperMill", 
                                          "other")) %>%
  ggplot(aes(free,rel)) + 
  geom_point(aes(color=set), 
             position=position_dodge(width=.7)) + 
  geom_errorbar(aes(x=free, ymin=CIl, ymax=CIu, color=set), 
                position=position_dodge(width=.7), width=.5) + 
  labs(x='Relative Number of Mentions', 
       y='Relative Number of Mentions') + 
  scale_color_manual('Type of Article', 
                     values = c("#2b83ba", "#ff8585")) +
  scale_y_continuous(limits = c(-2, 82)) +
  theme(legend.position = 'none', 
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        axis.text.y.left = element_blank()) +
  facet_grid(OriginalReason~name) ->
  p_creasons2

pp <- p_creasons1 + (p_creasons2 / plot_spacer() + 
                       plot_layout(heights = c(33.5,10)))
ggsave("two_column_plot.jpg", pp, width=12, height=7, bg = 'white')
pp + 
  plot_annotation(caption = 'Fig S17 (Article Fig. 5): Proportion of software mentions across 
    different levels of citation completeness per retraction reason, 
    separated by retracted and control articles. No Info: Neither the
    version, nor the developer of a software is provided; Incomplete 
    Info: Either version or developer are provided; Informal Citation: 
    Version and developer are provided; Formal citation: software 
    mention is accompanied by bibliographic citation (independent 
    from any associated information). Error bars indicate 95% CIs.') &
  theme(plot.caption = element_text(size=14))
```

### Per Reason (Open and Closed Source)

We also repeat this analysis for open and closed source software.

```{r, fig.width=11, fig.height=8.5}
df_reason_sampled %>%
  filter(!is.na(Software_ID)) %>%
  inner_join(software_enrichment, by=c(
    'Software_ID'='Software_ID', 
    'Software_Name'='Software_Name')) %>%
  group_by(set, source, OriginalReason, Paper_ID, Software_ID) %>%
  summarize(Version=any(Version), 
            Developer=any(Developer), 
            Citation=any(Citation), 
            .groups = 'drop_last') %>%
  distinct() %>%
  mutate(version_and_developer = Version & Developer & ! Citation, 
         no_citation_info = ! Version & ! Developer & ! Citation,
         version_or_developer = (Version & ! Developer & ! Citation) | 
           (Developer & ! Version & ! Citation),
         citation_p = Citation) %>% 
  dplyr::select(-Version, -Developer, -Citation) %>%
  rename(Citation=citation_p) %>%
  distinct() %>%
  group_by(set, source, OriginalReason) %>%
  summarize(`No Info` = sum(no_citation_info),
            `Incomplete Info` = sum(version_or_developer),
            `Formal Citation` = sum(Citation),
            `Informal Citation` = sum(version_and_developer),
            n = n(),
            .groups = 'drop_last') %>%
  pivot_longer(c(`No Info`, `Incomplete Info`, 
                 `Formal Citation`, `Informal Citation`)) %>%
  mutate(rel = value/n) %>%
  mutate(SEM=sqrt((rel * (1-rel))/n), 
         MoE = sqrt((rel * (1-rel))/n) * 1.96) %>%
  mutate(CIl = rel-MoE, CIu = rel + MoE) %>%
  group_by(name) %>%
  ungroup() %>%
  mutate(name=factor(name, levels=names)) %>%
  mutate(rel=rel*100, CIu=CIu*100, CIl=CIl*100) %>%
  mutate(source=ifelse(source, "Open", "Closed")) %>%
  mutate(set = ifelse('non-retracted'==set, 
                      "Control", 
                      "Retracted")) %>%
  mutate(OriginalReason = factor(OriginalReason, levels=reasons)) %>% 
  select(set, source, OriginalReason, name, rel, CIl, CIu) %>% 
  print(., n=128) ->
  df_creasons

filter(df_creasons, OriginalReason %in% c("Error", 
                                          "Investigation", 
                                          "Plagiarism", 
                                          "SelfPlagiarism")) %>%
  ggplot(aes(source,rel)) + 
  geom_point(aes(color=set), 
             position=position_dodge(width=.7)) + 
  geom_errorbar(aes(x=source, ymin=CIl, ymax=CIu, color=set), 
                position=position_dodge(width=.7), width=.5) + 
  labs(x='Relative Number of Mentions', 
       y='Relative Number of Mentions') + 
  scale_color_manual('Type of Article', 
                     values = c("#2b83ba", "#ff8585")) +
  theme(legend.position = 'top', 
        axis.title.x = element_blank(), axis.text.y = element_blank(), 
        axis.ticks.y.left = element_blank(),
        axis.text.y.right = element_text(hjust=.55),
        axis.title.y.right = element_blank(),
        plot.margin = unit(c(0,1,0,0), 'mm'),
        )  +
  facet_grid(OriginalReason~name, switch='y') +
    scale_y_continuous(
      limits = c(-2, 82), 
      breaks=c(0, 20, 40, 60, 80), 
      labels=c("0%", "20%", "40%", "60%", "80%"), 
      sec.axis = sec_axis(~., breaks=c(0, 20, 40, 60, 80), 
                          labels=c("0%", "20%", "40%", "60%", "80%"))) ->
    p_creasons1

filter(df_creasons, OriginalReason %in% c("Misconduct", 
                                          "PaperMill", 
                                          "other")) %>%
  ggplot(aes(source,rel)) + 
  geom_point(aes(color=set), 
             position=position_dodge(width=.7)) + 
  geom_errorbar(aes(x=source, ymin=CIl, ymax=CIu, color=set), 
                position=position_dodge(width=.7), width=.5) + 
  labs(x='Relative Number of Mentions', 
       y='Relative Number of Mentions') + 
  scale_color_manual('Type of Article', 
                     values = c("#2b83ba", "#ff8585")) +
  scale_y_continuous(limits = c(-2, 82)) +
  theme(legend.position = 'none', 
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        axis.text.y.left = element_blank()) +
  facet_grid(OriginalReason~name) ->
  p_creasons2


p_creasons1 + (p_creasons2 / plot_spacer() + 
                 plot_layout(heights = c(33.5,10)))  + 
  plot_annotation(
    caption = 'Fig S18: Proportion of software mentions across different levels 
    of citation completeness per retraction reason, separated by 
    retracted and control articles. No Info: Neither the version, 
    nor the developer of a software is provided; Incomplete Info: 
    Either version or developer are provided; Informal Citation: 
    Version and developer are provided; Formal citation: software 
    mention is accompanied by bibliographic citation (independent 
    from any associated information). Error bars indicate 95% CIs.') &
  theme(plot.caption = element_text(size=14))
```

```{r}
sessionInfo()
```
